---
title: "Analysis on the LNP-mRNA model"
author: "Yuezhe Li"
date: "4/19/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Model set

```{r pressure, echo=FALSE}
rm(list = ls())
gc()

setwd(dirname(rstudioapi::getSourceEditorContext()$path)) # set the working directory at the folder that contains the script

# load required packages
library(tidyverse)
library(mrgsolve)
library(gridExtra)
library(grid)
library(mrggsave)
library(sensitivity)
library(PKPDmisc)
library(mrgsim.parallel)

set.seed(88771) 
```

# Test simulation

```{r}
mod <- mread("model2") 

mod %>% param(dosing = 0.08, ktbg = 0) %>% mrgsim() %>% plot_sims(.f = LNP + mRNA + protein ~ time/(60*60*24))
```

# Global sensitivity analysis

## mRNA

use mRNA production as the readout

```{r}
mod <- mread("model2") %>% param(dosing = 0.08, ktbg = 0)

# create sampling method
gen_samples <- function(n, l, which = names(l), 
                        factor = c(0.01,100), N = NULL) { # NEW ARGUMENT: N, the absolute sampling number per parameter
  
  vars <- tidyselect::vars_select(names(l), !!(enquo(which)))
  
  l <- as.list(l)[vars]
  
  l <- lapply(l, function(x) x*factor )
  
  if(is.numeric(N)) { # NEW
    n <- N*2  
  } else {
    n <- length(l)*n*2
  }
  
  df <- as.data.frame(l)
  
  len <- length(df)
  
  X <- matrix(ncol=len, nrow=n)
  
  colnames(X) <- names(df)
  
  Y <- X
  
  for(i in seq(len)){
    r <- exp(runif(n, log(df[1,i]), log(df[2,i])))
    X[,i] <- r
    r <- exp(runif(n, log(df[1,i]), log(df[2,i])))
    Y[,i] <- r
  }
  
  return(list(x1 = as.data.frame(X), x2 = as.data.frame(Y)))
}


# set up simulation parameters for sobol2007 function
simulationboot = 200 # number of simulation time in sobol2007
sampleperparam = 100 # sampling size per parameter, for sobol2007

setpk <- c('kw', 'k12', 'k21')
setpd <- c('ka', 'ke','de' ,'kl', 'dmRNA')

N <- sampleperparam * length(c(setpk,setpd))

# set different variation in each group
sets <- list(
  list(which = setpk, factor = c(0.5, 1.5), N = N, l = param(mod)), 
  list(which = setpd, factor = c(0.01,100), N = N, l = param(mod))
)

samps <- map(sets, do.call, what = gen_samples)
samp <- map(c(1,2), ~ map_dfc(samps, .x))


batch_mRNA <- function(x) {
    mod %>% 
    idata_set(x) %>%
    mrgsim(obsonly = TRUE) %>% 
    group_by(ID) %>% 
    summarise(pexp = auc_partial(time, mRNA)) %>% 
    pull(pexp)
}

# warning: the following line takes several minutes to run
# pglobal = sobol2007(batch_mRNA, X1=samp[[1]], X2=samp[[2]], nboot=simulationboot)
# plot(pglobal)
# lines(c(0,11), c(0.05, 0.05), type="l", lty = 2)
# save a minimal set of the Sobol analysis data
# that is sufficient for visualization
sobolx <- list(S = pglobal$S , T = pglobal$T)
saveRDS(sobolx, file = "data/Sobol_mRNA.rds")
```

Plot the global sensitivity analysis, mRNA

```{r}
x <- readRDS("data/Sobol_mRNA.rds")

setpk <- c('kw', 'k12', 'k21')
setpd <- c('ka', 'ke','de' ,'kl', 'dmRNA')

### plot
globSens_mRNA <- tibble(Parameter = c(setpk,setpd),
                   main = x$S$original,
                   main_lo = x$S$'min. c.i.',
                   main_hi = x$S$'max. c.i.',
                   total = x$T$original,
                   total_lo = x$T$'min. c.i.',
                   total_hi = x$T$'max. c.i.') %>%
  gather(effect, Index, -Parameter, -main_lo, -main_hi, -total_lo, -total_hi) %>%
  mutate(lo = ifelse(effect == "main", main_lo, total_lo),
         hi = ifelse(effect == "main", main_hi, total_hi),
         Effect = factor(effect))

fig_sens_glob_mRNA <- ggplot(data=globSens_mRNA, aes(x=Parameter, y=Index, group=Effect, col=Effect)) +
  geom_point(position = position_dodge(width=0.3)) +
  geom_errorbar(aes(ymax=hi, ymin=lo), position=position_dodge(width=0.3), width=0) +
  theme_bw() +
  geom_hline(aes(yintercept=0.05), lty=2) +
  theme(legend.title = element_text(size = 15), legend.text=element_text(size=12)) + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size = 12)) + 
  ggtitle('Global sensitivity analysis, use mRNA as a read out')


print(fig_sens_glob_mRNA)

ggsave(filename = 'img/GlobalSens_mRNA.png', plot = fig_sens_glob_mRNA, height = 3, width = 8, units = c('in'), dpi = 300)
```

## Protein

Use protein expression as a readout. Focus on the mRNA-protein dynamics.

```{r}
mod <- mread("model2") %>% param(dosing = 0.08, ktbg = 0)

# set up simulation parameters for sobol2007 function
simulationboot = 800 # number of simulation time in sobol2007
sampleperparam = 250 # sampling size per parameter, for sobol2007

setprotein <- c('ka', 'ke','de' ,'kl', 'dmRNA', 'kt', 'dUGTc')

N <- sampleperparam * length(setprotein)

# set different variation in each group
setp <- list(
  list(which = setprotein, factor = c(0.05,20), N = N, l = param(mod))
)

samps <- map(setp, do.call, what = gen_samples)
samp <- map(c(1,2), ~ map_dfc(samps, .x))


batch_protein <- function(x) {
    mod %>% 
    idata_set(x) %>%
    mrgsim(obsonly = TRUE) %>% 
    group_by(ID) %>% 
    summarise(pexp = auc_partial(time, protein)) %>% 
    pull(pexp)
}

# warning: the following line takes several minutes to run
pglobal = sobol2007(batch_protein, X1=samp[[1]], X2=samp[[2]], nboot=simulationboot)
plot(pglobal)
# lines(c(0,11), c(0.05, 0.05), type="l", lty = 2)
# save a minimal set of the Sobol analysis data
# that is sufficient for visualization
sobolx <- list(S = pglobal$S , T = pglobal$T)
saveRDS(sobolx, file = "data/Sobol_protein.rds")
```
Plot the global sensitivity analysis, protein

```{r}
x <- readRDS("data/Sobol_protein.rds")

setprotein <- c('ka', 'ke','de' ,'kl', 'dmRNA', 'kt', 'dUGTc')

### plot
globSens_protein <- tibble(Parameter = setprotein,
                   main = x$S$original,
                   main_lo = x$S$'min. c.i.',
                   main_hi = x$S$'max. c.i.',
                   total = x$T$original,
                   total_lo = x$T$'min. c.i.',
                   total_hi = x$T$'max. c.i.') %>%
  gather(effect, Index, -Parameter, -main_lo, -main_hi, -total_lo, -total_hi) %>%
  mutate(lo = ifelse(effect == "main", main_lo, total_lo),
         hi = ifelse(effect == "main", main_hi, total_hi),
         Effect = factor(effect))

fig_sens_glob_protein <- ggplot(data=globSens_protein, aes(x=Parameter, y=Index, group=Effect, col=Effect)) +
  geom_point(position = position_dodge(width=0.3)) +
  geom_errorbar(aes(ymax=hi, ymin=lo), position=position_dodge(width=0.3), width=0) +
  theme_bw() +
  geom_hline(aes(yintercept=0.05), lty=2) +
  theme(legend.title = element_text(size = 15), legend.text=element_text(size=12)) + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size = 12)) + 
  ggtitle('Global sensitivity analysis, use protein as a read out')


print(fig_sens_glob_protein)

ggsave(filename = 'img/GlobalSens_protein.png', plot = fig_sens_glob_protein, height = 3, width = 8, units = c('in'), dpi = 300)
```

# Local Sensitivity Analysis

## LNP attachment rate

```{r}
idata <- expand.idata(ka = c(1E-6, 1E-5, 5E-5, 1E-4))

out <- mod <- mread("model2") %>% param(dosing = 0.08, ktbg = 0) %>% 
  idata_set(idata) %>% mrgsim(end = 60*60*24*7) %>% as.tibble() %>%
  mutate(time =  time / (60*60*24) ) 

ka_protein <- ggplot(data = out, aes(x = time, y = protein, col = as.factor(ID) ) ) + geom_line() + 
  scale_colour_discrete( breaks=c(1,2,3,4), labels=c("ka=1E-6", "ka=1E-5", "ka=5E-5", "ka=1E-4")) + 
  theme_bw() + theme(legend.title = element_blank(), legend.position="bottom") + 
  labs(x = "time (day)", y = "protein (nmol)", title = "Parameter scan, LNP attachment rate")


ka_mrna <- ggplot(data = out, aes(x = time, y = mRNA, col = as.factor(ID) ) ) + geom_line() + 
  scale_colour_discrete( breaks=c(1,2,3,4), labels=c("ka=1E-6", "ka=1E-5", "ka=5E-5", "ka=1E-4")) + 
  theme_bw() + theme(legend.title = element_blank(), legend.position="bottom") + 
  labs(x = "time (day)", y = "mRNA (nmol)", title = "Parameter scan, LNP attachment rate") 

png('img/localsens_ka.png', width = 12, height = 3, unit = "in", res = 300)
grid.arrange(ka_mrna, ka_protein, ncol = 2)
dev.off()
```


## LNP degredation rate in endosome

```{r}
idata <- expand.idata(de = c(1E-5, 1E-4, 1E-3))

out <- mod <- mread("model2") %>% param(dosing = 0.08, ktbg = 0) %>% 
  idata_set(idata) %>% mrgsim(end = 60*60*24*7) %>% as.tibble() %>%
  mutate(time =  time / (60*60*24) ) 

de_protein <- ggplot(data = out, aes(x = time, y = protein, col = as.factor(ID) ) ) + geom_line() + 
  scale_colour_discrete( breaks=c(1,2,3), labels=c("de=1E-5", "de=1E-4", "de=1E-3")) + 
  theme_bw() + theme(legend.title = element_blank(), legend.position="bottom") + 
  labs(x = "time (day)", y = "protein (nmol)", title = "Parameter scan, LNP endosomal degradation rate")


de_mrna <- ggplot(data = out, aes(x = time, y = mRNA, col = as.factor(ID) ) ) + geom_line() + 
  scale_colour_discrete( breaks=c(1,2,3), labels=c("de=1E-5", "de=1E-4", "de=1E-3")) + 
  theme_bw() + theme(legend.title = element_blank(), legend.position="bottom") + 
  labs(x = "time (day)", y = "mRNA (nmol)", title = "Parameter scan, LNP endosomal degradation rate") 

png('img/localsens_de.png', width = 12, height = 3, unit = "in", res = 300)
grid.arrange(de_mrna, de_protein, ncol = 2)
dev.off()

```


## change protein translation rate
```{r}
idata <- expand.idata(kt = c(10, 17, 34, 100))

out <- mod <- mread("model2") %>% param(dosing = 0.08, ktbg = 0) %>% idata_set(idata) %>% mrgsim(end = 60*60*24*7) %>% as.tibble() %>%
  mutate(time =  time / (60*60*24) ) 

ktscan <- ggplot(data = out, aes(x = time, y = protein, col = as.factor(ID) ) ) + geom_line() + theme_bw() + 
  scale_colour_discrete( breaks=c(1,2,3,4), labels=c("kt=10", "kt=17", "kt=34", "kt=100")) + theme(legend.title = element_blank()) + 
  labs(x = "time (day)", y = "protein (nmol)", title = "Parameter scan, translation rate") 


ggsave(filename = 'img/localsens_protein_kt.png', plot = ktscan, height = 3, width = 10, units = c('in'), dpi = 300)

```

## protein degredation rate

```{r}
idata <- expand.idata(dUGTc = c(5E-7, 1E-6, 5E-6, 2E-5))

out <- mod <- mread("model2") %>% param(dosing = 0.08, ktbg = 0) %>% idata_set(idata) %>% mrgsim(end = 60*60*24*21) %>% as.tibble() %>%
  mutate(time =  time / (60*60*24) ) 

dUGTc_scan <- ggplot(data = out, aes(x = time, y = protein, col = as.factor(ID) ) ) + geom_line() + theme_bw() + 
  scale_colour_discrete( breaks=c(1,2,3,4), labels=c("dUGTc=5E-7", "dUGTc=1E-6", "dUGTc=5E-6", "dUGTc=2E-5")) + 
  theme(legend.title = element_blank()) + 
  labs(x = "time (day)", y = "protein (nmol)", title = "Parameter scan, protein degradation rate") 


ggsave(filename = 'img/localsens_protein_dUTGc.png', plot = dUGTc_scan, height = 3, width = 10, units = c('in'), dpi = 300)

```
